A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm

### 2014-10-28 paint-contour-tree.tex

Here we present the first contour tree algorithm which takes into account the shape of the contour tree, 
and in doing so yields instance optimal results.
Specifically, we prove that for any partition of the contour tree into a set of descending paths, $P$, 
any comparison based algorithm requires $\sum_{p\in P} |p|\log|p|$ time.  Moreover, we show there   
exists some partition $P'$ such that the running time of our algorithm is 
$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is total complexity of the input manifold 
(and hence this is optimal on all instances requiring $\Omega(N\alpha(N))$ time).
In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, 
while previous algorithms ran in $O(N+t\log t)$ (where $t$ is the size of the contour tree).
%In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, yet still 
%matches the previous worst case $O(N+n\log n)$ bound when the tree is long and skinny.

@@

\subsubsection{Worst Case Alternatives}
In worst case analysis an algorithm's running time is judged in terms of its worst 
case behavior over all inputs of a given size.  While there are benefits to such a 
pessimistic view, it fails to capture the whole story.  As such, many alternatives 
have been considered.  For example, one can make assumptions on the distribution of 
inputs, which has led to a number interesting results on average case and probabilistic 
analysis, and also smoothed analysis \cite{st-saa-04}.  However, even without making 
assumptions on the distribution of the input, there is still more to the story.
For example, in the study of parametrized complexity, one takes into account other 
parameters determined by the input other than just its size (for example the maximum 
degree in a graph or whether it is planar).

@@

The previous section allows us to restrict attention to extremum dominant manifolds.
We will orient so that the extremum in question is always a \emph{minimum}.
We will fix such a simplicial complex $\MM$, with the dominant minimum $m^*$. 
We observe that the contours trees of minimum dominant manifolds have a relatively simple description.
This description is directly inspired by (and proven using) the \emph{join tree} of~\cite{csa-cctad-00}.
We first explain what these trees are.

For vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
keeping vertices $u$ such that $f(u) > f(v)$. Analogously, define $\MM^-_v$. Note that $\MM^+$ may contain
numerous connected components. The following definition of join and split trees were given by~\cite{csa-cctad-00}.
Conventially, all edges are directed from higher to lower function value.

\begin{definition} \label{join} The join tree $\cJ(\MM)$ of $\MM$ is built on vertex set $V(\MM)$.
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in a connected component of $\MM^+_v$
\emph{and} $v$ is connected to this component (in $\MM$). The split tree $\cS(\MM)$ is obtained
by looking at $\MM^-_v$ (or alternately, by taking the join tree of the inversion of $\MM$). 
\end{definition}

Note that unlike the contour tree, all vertices in $\MM$ are present in $\cJ(\MM)$ and $\cS(\MM)$. 
Some basic facts about these trees. All outdegrees in $\cJ(\MM)$ and all indegrees in $\cS(\MM)$ are at most $1$.
Indegree $2$ vertices in $\cJ(\MM)$ are joins, and outdegree $2$ vertices in $\cS(\MM)$ are splits.
Leaves in $\cJ(\MM)$ are maxima, and leaves in $\cS(\MM)$ are minima.

The key observation is that $\cS(\MM)$ is trivial for minimum dominant $\MM$.

@@

Now we come to the main algorithmic contribution. We design a new algorithm for computing contour
trees of a minimum dominant simplicial complex $\MM$. As part of the initialization, we determine
all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
vertices.

{\bf Painting:} The central tool is a notion 
of \emph{painting} of $\MM$. Initially associate a color with each maximum. Denoting the set of maxima by $X$,
this is an arbitrary bijection $\chi:X \mapsto [|X|]$. Imagine there being a large
can of paint of color $\chi(x)$ at maxima $x$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of facets.} Furthermore, paints
do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
works purely on the 1-skeleton of $\MM$, which is just a graph.
%
\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ be denoted $G(\MM) = (\cV,E)$.
A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
 Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
	is well-defined to be the unique maximum with the same color as $e$.)

>>>>

Here we present the first contour tree algorithm which takes into account the shape of the contour tree, 
and in doing so yields instance optimal results.
Specifically, we prove that for any partition of the contour tree into a set of descending paths, $P$, 
any comparison based algorithm requires $\sum_{p\in P} |p|\log|p|$ time.  Moreover, we show there   
exists some partition $P'$ such that the running time of our algorithm is 
$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is the total complexity of the input manifold 
(and hence this is optimal on all instances requiring $\Omega(N\alpha(N))$ time).
In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, 
while previous algorithms ran in $O(N+t\log t)$ (where $t$ is the size of the contour tree).
%In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, yet still 
%matches the previous worst case $O(N+n\log n)$ bound when the tree is long and skinny.

@@

\subsubsection{Worst Case Alternatives}
In worst case analysis an algorithm's running time is judged in terms of its worst 
case behavior over all inputs of a given size.  While there are benefits to this viewpoint, 
such as robustness,  it fails to capture the whole story.  As such, many alternatives 
have been considered.  For example, one can make assumptions on the distribution of 
inputs, which has led to a number interesting results on average case and probabilistic 
analysis, and also smoothed analysis \cite{st-saa-04}.  However, even without making 
assumptions on the distribution of the input, there is still more to the story.
For example, in the study of parametrized complexity, one takes into account other 
parameters determined by the input other than just its size (for example the maximum 
degree in a graph or whether it is planar).

@@

The previous section allows us to restrict attention to extremum dominant manifolds.
We will orient so that the extremum in question is always a \emph{minimum}.
We will fix such a simplicial complex $\MM$, with the dominant minimum $m^*$. 
The set of non-dominant minima is denoted by $M$.
For vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
keeping vertices $u$ such that $f(u) > f(v)$. Analogously, define $\MM^-_v$. Note that $\MM^+_v$ may contain
numerous connected components.

The main theorem of this section asserts that the contours trees of minimum dominant manifolds have a simple description.
This description is based on \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-00}.
Conventially, all edges are directed from higher to lower function value.
We first explain what these trees are.

\begin{definition} \label{join} The join tree $\cJ(\MM)$ of $\MM$ is built on vertex set $V(\MM)$.
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in a connected component of $\MM^+_v$
\emph{and} $v$ is connected to this component (in $\MM$). The split tree $\cS(\MM)$ is obtained
by looking at $\MM^-_v$ (or alternately, by taking the join tree of the inversion of $\MM$).

The critical join tree $\jc(\MM)$ is built on $V(\MM) \setminus M$. 
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V(\MM) \setminus M$ in a connected component of $\MM^+_v$
and $v$ is connected to this component (in $\MM$). 
\end{definition}

Some basic facts about these trees. 
%Note that unlike the contour tree, all vertices in $\MM$ are present in $\cJ(\MM)$ and $\cS(\MM)$. 
All outdegrees in $\cJ(\MM)$ and all indegrees in $\cS(\MM)$ are at most $1$.
Indegree $2$ vertices in $\cJ(\MM)$ are joins, and outdegree $2$ vertices in $\cS(\MM)$ are splits.
Leaves in $\cJ(\MM)$ are maxima, and leaves in $\cS(\MM)$ are minima.
The key observation is that $\cS(\MM)$ is trivial for minimum dominant $\MM$.

@@

The main algorithmic contribution is a new algorithm for computing join trees of any triangulated simplicial
complex $\MM$.

As part of the initialization, we determine
all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
vertices.

{\bf Painting:} The central tool is a notion 
of \emph{painting} of $\MM$. Initially associate a color with each maximum. Imagine there being a large
can of paint of a distinct color at each maxima $x$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of facets.} Furthermore, paints
do not mix, so every edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
works purely on the 1-skeleton of $\MM$, which is just a graph.
%
\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
 Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
	is well-defined to be the unique maximum with the same color as $e$.)
